Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-01-31 ms-3/ms.tex

\subsection*{Application to empirical data}
We applied our framework to COVID-19 case data for the United States at the administrative level of counties, compiled by The New York Times, based on reports from state and local health agencies between Jan 4, 2020, and March 18, 2023~\citep{nytimes_covid19_2021}, and using county population sizes estimated for 2021 from the United States Census Bureau~\citep{us_census_county_estimates_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differencing to compute new cases. Occasionally, reported cumulative case counts were not monotonically increasing due to corrections posted by local agencies as they resolved incoming data. 
As a result, approximately 0.24\% of estimated new cases across all counties in the dataset were negative and these were set to zero. 
For each county, we analyzed overlapping 16-week windows, shifting one week (i.e., one timestep) at a time. 
Within each window, the framework estimated the dispersion parameter ($\theta$) using a natural spline with three degrees of freedom to model the broad-scale trend in incidence. 
Outputs included estimated dispersion parameters ($\theta_1$, $\theta_2$, for the left and right halves of each window, and $\theta$ for the entire window), likelihood ratio test statistics, $p$-values for changes in dispersion at the midpoint of the window, and flags for boundary conditions such as failure to reject Poisson-like dispersion or collapse to extreme overdispersion.

\section*{Results and Discussion}
Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}). The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million. Across this range, the method produced accurate estimates for $\theta$ within $10^{-2} \leq \theta \leq 10^2$, encompassing all operationally relevant values for COVID-19 incidence data and many other infectious diseases. Lower values would concentrate the probability mass function (PMF) for cases almost entirely on 0, while higher values effectively correspond to a Poisson distribution.

>>>>

\subsection*{Application to empirical data}
We applied our framework to COVID-19 case data for the United States at the administrative level of counties, compiled by The New York Times, based on reports from state and local health agencies between Jan 4, 2020, and March 18, 2023~\citep{nytimes_covid19_2021}, and using county population sizes estimated for 2021 from the United States Census Bureau~\citep{us_census_county_estimates_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differencing to compute new cases. Occasionally, reported cumulative case counts were not monotonically increasing due to corrections posted by local agencies as they resolved incoming data. 
As a result, approximately 0.24\% of estimated new cases across all counties in the dataset were negative and these were set to zero. 
For each county, we analyzed overlapping 16-week windows, shifting one week (i.e., one timestep) at a time. 
Within each window, the framework estimated the dispersion parameter ($\theta$) using a natural spline with three degrees of freedom for each half of the window to model the broad-scale trend in incidence. 
Outputs included estimated dispersion parameters ($\theta_1$, $\theta_2$, for the left and right halves of each window, and $\theta$ for the entire window), likelihood ratio test statistics, $p$-values for changes in dispersion at the midpoint of the window, and flags for boundary conditions such as failure to reject Poisson-like dispersion or collapse to extreme overdispersion.

\section*{Results and Discussion}

Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}). The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million. Across this range, the method produced accurate estimates for $\theta$ within $10^{-2} \leq \theta \leq 10^2$, encompassing all operationally relevant values for COVID-19 incidence data and many other infectious diseases. Lower values would concentrate the probability mass function (PMF) for cases almost entirely on 0, while higher values effectively correspond to a Poisson distribution.
